\DocumentMetadata
  {
    lang=en-US,
    pdfversion=2.0,
    pdfstandard=ua-2,
    testphase={phase-III,title,math,table,firstaid}
  }
\documentclass{article}
\usepackage{poetry}

\title{poetry tagging test}

\begin{document}

\begin{poem}
Let me not to the marriage of true minds \\
\hin Admit impediments. Love is not love \\
Which alters when it alteration finds, \\
\hin Or bends with the remover to remove: \\!
O, no! it is an ever-fixed mark, \\
\hin That looks on tempests and is never shaken; \\
It is the star to every wandering bark, \\
\hin Whose worth's unknown, although his height be taken. \\-
\end{poem}

\bgroup
\centerpoemon{The ship has weather'd every rack, the prize we sought is won}
\begin{poem}
O captain! my captain! The fearful trip is done; \\
The ship has weather'd every rack, the prize we sought is won;\\-
\end{poem}
\egroup

\bgroup
\centerpoemon{Alas, poor Yorick! I knew him. Horatio!}
\settowidth{\poemhinwd}{Alas, poor Yorick! I knew him. }
\begin{poem}
Alas, poor Yorick! I knew him.\\
\hin Horatio! \\
\setlength{\poemhinwd}{2em}
A fellow of infinite jest, of most excellent fancy. \\
\hin He hath borne me on his back a thousand times, \\
and now, how abhorred in my imagination it is! \\
My gorge rises at it. Here hung those lips...\\-
\end{poem}
\egroup

\bgroup
\setcounter{poemindentevery}{3}
\setcounter{poemlinenumsevery}{2}
\begin{poem}
Once upon a midnight dreary, while I pondered, weak and weary,\\
Over many a quaint and curious volume of forgotten lore---\\
While I nodded, nearly napping, suddenly there came a tapping,\\
As of someone gently rapping, rapping at my chamber door.\\
``'Tis some visitor,'' I muttered, ``tapping at my chamber door---\\
Only this and nothing more.''\\-
\end{poem}
\egroup

\bgroup
\poemlinenumsfalse
\begin{poem}
O captain! my captain! The fearful trip is done; \\
\poemlinenumstrue
The ship has weather’d every rack, the prize we sought is won;\\
The port is near, the bells I hear, the people all exulting, \\
While follow eyes the steady keel, the vessel grim and daring; \\-
\end{poem}
\egroup

\bgroup
\poemlinenumrighttrue
\setcounter{poemlinenumsevery}{1}
\titleauthorpoem{\poemblanktitle}{\poemblankauthor}
\begin{poem}
O captain! my captain! The fearful trip is done; \\
\poemlinenumrightfalse
The ship has weather’d every rack, the prize we sought is won;\\
The port is near, the bells I hear, the people all exulting, \\
While follow eyes the steady keel, the vessel grim and daring; \\-
\end{poem}
\egroup

\end{document}
