\DocumentMetadata
  {
    lang=en-US,
    pdfversion=2.0,
    pdfstandard=ua-2,
    tagging=on,
    %uncompress,
  }
\documentclass{article}
\usepackage[top=2cm,bottom=2cm,left=1cm,right=1cm]{geometry}
\usepackage{hyperref,enumext}
\hypersetup
  {
    colorlinks = true,
    pdftitle   = {Test tagged PDF for enumext package},
  }
\begin{document}
\section{only enumext}
Some text here

\begin{enumext}[labelwidth=1cm,labelsep=0.25cm,align=left, wrap-label=\textbf{(#1}, label=\arabic*.)]
  \item This text is in the first level.\label{A}
    \begin{enumext}
      \item This text is in the second level.\footnote{foot note in enumext}
        \begin{enumext}
          \item This text is in the third level.
            \begin{enumext}
              \item This text is in the fourth level.\footnotemark[20]
            \end{enumext}
        \end{enumext}
    \end{enumext}
  \item[X] This text is in the first level.
  \item* This text is in the first level.
\end{enumext}

Some ref here \ref{A}\footnotetext[20]{foot note in enumext [20]}

\section{only enumext*}

Some text here

\begin{enumext*}[columns=2,labelwidth=1cm,labelsep=0.25cm,align=left, wrap-label=\textbf{#1)}, label=(\arabic*.]
  \item This text is in the first level.\label{B}
  \item[X] This text is in the first level.\footnotemark[30]
  \item* This text is in the first level.\footnote{foot note in enumext*}
  \item This text is in the first level.
\end{enumext*}

Some ref here \ref{B}\footnotetext[30]{foot note in enumext* [30]}

\section{enumext nested in enumext*}

Some text here

\begin{enumext*}
  \item This text is in the first level.\label{C}
  \item \begin{enumext}[base-fix]
              \item This text is in nested enumext.\label{D}\footnote{foot note in neseted enumext}
              \item This text is in nested enumext.\footnotemark[40]
         \end{enumext}
  \item This text is in the first level.
\end{enumext*}

Some refs here \ref{C} and \ref{D} \footnotetext[40]{foot note in neseted enumext [40]}

\section{enumext* nested in enumext}

\begin{enumext}
  \item This text is in the first level. \label{E}
  \item \begin{enumext*}[label=\roman*.]
              \item This text is in nested enumext*.\label{F}\footnotemark[50]
              \item This text is in nested enumext*.\footnote{foot note in neseted enumext*}
         \end{enumext*}
  \item This text is in the first level.
\end{enumext}

Some refs here \ref{E} and \ref{F} \footnotetext[50]{foot note in neseted enumext* [50]}

\section{Custom ref}

\begin{enumext}[ref={enumext-level-\arabic*}]
  \item This text is in the first level. \label{G}
  \item \begin{enumext*}[label=\roman*., ref={nested-\arabic{enumXi}.\roman*}]
              \item This text is in nested enumext*.\label{H}
              \item This text is in nested enumext*.
         \end{enumext*}
  \item This text is in the first level.
\end{enumext}

Some refs here \ref{G} and \ref{H}
\end{document}
