% !Mode:: "TeX:DE:UTF-8:Main"
%
% If it should be processed without tagging change \iftrue to \iffalse.
%

\iftrue

\DocumentMetadata{
   tagging=on,
   pdfversion=2.0,
   pdfstandard=a-4,
   pdfstandard=ua-2
}

\fi

\documentclass[final]{ltugboat}
\let\chapter\undefined
\providecommand\email[1]{\href{mailto:#1}{{\ttfamily #1}}}

\usepackage{microtype,csquotes,graphicx}
\usepackage[hidelinks]{hyperref}

\providecommand\pkg[1]{\texttt{#1}}
\providecommand\env[1]{\texttt{#1}}
\providecommand\Dash{\,---\,}
\providecommand\PDF{PDF}
\providecommand\pdfTeX{pdf\TeX}
\providecommand\LuaTeX{Lua\TeX}
\providecommand\upTeX{up\TeX}


%Actualtext for the Logos
\ExplSyntaxOn
\IfDocumentMetadataTF{
\cs_new:Npn \AddActualTextToCommand #1#2
{
  \AddToHook{cmd/#1/before}
    {
      \mode_leave_vertical:
      \tag_mc_end_push:
      \tag_struct_begin:n{tag=Span,actualtext=#2}
      \tag_mc_begin:n{}
      \tag_stop:n{#1}
    }
  \AddToHook{cmd/#1/after}
   {
     \tag_start:n{#1}
     \tag_mc_end:
     \tag_struct_end:
     \tag_mc_begin_pop:n{}
   }
}
\AddActualTextToCommand{pdfTeX}{PDFTeX}
\AddActualTextToCommand{upTeX}{upTeX}
\AddActualTextToCommand{XeTeX}{XeTeX}
\AddActualTextToCommand{LuaTeX}{LuaTeX}
}{}
\ExplSyntaxOff

%% Front Matter
%%
% Regular title as in the article class.
%
\title{Enhancing \LaTeX{} to automatically produce tagged and accessible \PDF\thanks{This
article is a fully tagged and accessible \PDF\ produced by the project software it describes.}}

\author{Frank Mittelbach}
\address{Mainz, Germany}
\netaddress{https://www.latex-project.org}
%\personalURL{}

\author{Ulrike Fischer}
\address{Bonn, Germany}
\netaddress{https://www.latex-project.org}



\begin{document}
 \maketitle

% \looseness=-1
 \noindent
 This paper was initially presented at the 5th International Workshop on
\enquote{Digitization and E-Inclusion in Mathematics and Science 2024} (\acro{DEIMS}~2024)
15--17 February 2024, at Nihon University, Tokyo, Japan.
This version contains some minor updates.

%The workshop website contains the full program and material on all presentations~\cite{deims2024}.
The complete program and  material on all presentations can be found at the workshop website~\cite{deims2024}.
%
A video of the talk and the demonstration is available at \url{https://youtu.be/7FnZv5FhmRg&?t=9869}.

\begin{abstract}
  At the \acro{TUG} 2020 online conference the \LaTeX{} Project Team
  announced the start of a multi-year project to enhance \LaTeX{} so
  that it will fully and naturally support the creation of structured
  document formats, in particular the \enquote{tagged \PDF} format as
  required by accessibility standards such as \acro{PDF/UA}.

  In this talk we present the current achievements of this
  project\footnote{This project is carried out by a small number of
  developers. Besides the authors, the following individuals from
  the \LaTeX{} Project Team are actively involved:
%
   Chris Rowley,
   David Carlisle,
   Joseph Wright,
   Marcel Krüger, and
   Phelype Oleinik.

  We also wish to acknowledge the contributions from various
  members of the \TeX{} community and beyond; these have been made
  through comments and suggestions, and more recently through the
  testing of the new functionality made available in the form of
  prototype implementations. Without such feedback it would be
  difficult to finish this project with satisfactory results.}
%
  and the issues we encountered along the way. We also outline open
  areas of research and the future steps that we shall take to
  automatically produce well-tagged \PDF\ that supports accessible
  standards (in particular, the recently finalized \acro{PDF/UA-2}) as well
  as general reuse and further conversions. This will be achieved by
  embedding in the \PDF\ a comprehensive description of the document
  structure.
\end{abstract}

% The table of contents below is added for your convenience. Please do not use
% the table of contents if you are preparing your paper for publication in the
% EPiC Series or Kalpa Publications series

\setcounter{tocdepth}{2}
{\small\tableofcontents}


%------------------------------------------------------------------------------

\section{General overview}

For over 30 years now, the \LaTeX{} system has been used, widely and
successfully, for document production in the \acro{STEM} world and also in
other places where high-quality output is required; but until recently
its focus was solely on page-oriented output for print (on paper) or
as paged output using the \PDF\ format.  Therefore, the structural
information about the document that was present in the \LaTeX{} source
did not get incorporated into the \PDF\ output.  Rather, this
information was discarded as soon as possible during the processing;
this was necessary so as to conserve the limited computer resources
(memory and storage) that were typically available at that time (when
the core of the \LaTeX{} processing model was first designed).

As long as the intention is only to print a document on a physical
medium, then this is all that is required. However, for quite a while
now other uses of documents have been increasing in importance so that
nowadays many documents are never printed, or printed only as a
secondary consideration.

Coming into the 21st century, for many reasons great interest has
arisen in the production of \PDF\ documents that
are \enquote{accessible}, in the sense that they contain information
to assist screen reading software, etc., and, more formally, that they
adhere to the \acro{PDF/UA} (Universal Accessibility)
standard~\cite{ISO14289-1:2014,ISO14289-2:2024}, which is explained
further in~\cite{pdf-ua-nutshell}.

At present, all methods for producing such \enquote{accessible \PDF{}s},
including the use of \LaTeX{}, require extensive manual
labor\footnote{If not using the already existing code extensions
to \LaTeX{} provided by the project.}  during either the preparation
of the source or the post-processing of the \PDF\ (maybe even at both
stages); and these labors often have to be repeated after making even
minimal changes to the (\LaTeX{} or other) source.


\subsection{The goals of the multi-year \texorpdfstring{\\}{}\enquote{\LaTeX{} Tagged \PDF} project}


The main goal of the project is to enhance \LaTeX{} so that it can
\emph{automatically} produce tagged \PDF\ without the need to add additional
data or commands to the \LaTeX{} source, or to do any of the
post-processing work necessary in other workflows.

If it remains necessary to alter substantially, or to extend, each
individual document in order to provide tagged \PDF\ that conforms to
some accessibility standard, then we shall see very few document
authors willing to go through the pain of making such additions
(unless they are forced to). It is therefore of utmost importance that
the generation of tagged \PDF\ be done essentially behind the scenes,
with the only cost to the authors being a somewhat longer compilation
time.

Another important aim is to make already existing documents accessible
by simply recompiling them without the need to alter the source in any
substantial way.\footnote{Of course, required data that is not part of
the document source (such as alternative text for figures or
additional metadata) will need to be manually added, so as to ensure
that the document is compliant with \acro{PDF/UA}. But even if this work is
not undertaken, the fact that the document gets automatically tagged
will mean that it can be easily navigated and consumed in ways that
were impossible before.}

The project will support the \PDF~2.0
standard~\cite{pdfspec-iso32000-2_2020} (with the very widely
supported \PDF~1.7 as a fallback solution), because the \PDF~2.0
standard offers a more comprehensive tag set, and it supports
associated files and many other important features; its use is also a
requirement of the new \acro{PDF/UA-2} standard~\cite{ISO14289-2:2024}.
{\hfuzz=1.2pt\par}

Unfortunately, even though \PDF~2.0 has already existed for six years,
it has yet to be adopted for industry solutions; e.g., most viewers
and other applications are still incapable of making correct use of
the new \PDF~2.0 features.  This is largely a chicken-and-egg problem:
because nobody produced 2.0 files, no application was specifically
extended to enable processing such files; and due to the fact that no
viewer could handle such files, the developers of \PDF\ writers saw no
need to invest in the technology to produce \PDF~2.0 files.

As a result, \LaTeX{} is one of the first authoring applications that
can produce \PDF~2.0 files automatically and in large quantities. In
particular, \LaTeX{} is capable of producing documents compliant with
\acro{PDF/UA-2}, the new standard for Universal
Accessibility~\cite{ISO14289-2:2024} that was finalized in 2023 and
will be officially released in early 2024.\footnote{At the moment we
can only claim that the project software is capable of producing
documents that comply with the latest draft of the imminent \acro{PDF/UA-2}
standard.}
%
No doubt other suppliers will follow our lead when there is sufficient
demand for the production and processing of \PDF~2.0 and \acro{PDF/UA-2}
conformant files.

The document entitled \enquote{\LaTeX{} Tagged \PDF\ Feasibility
Evaluation}~\cite{feasibility-study}, available from the \LaTeX{}
Project website~\cite{latex-project-website}, explains in detail both
the project goals and the tasks that need to be undertaken, concluding
with the project plan that is currently being executed.

For the time being the project will focus primarily on \PDF\ output
(generated either directly by the \TeX{} engine or through a \DVI-based
workﬂow). However, as a bonus outcome of the design approach, the
implemented solution will make it easy to add other such output
formats to the workﬂow by simply replacing the output (backend)
module.  Instead of \PDF\ output, \HTML5 or some other format can thus be
written. As of now, such alternative backends are not part of the
project coverage, but once \LaTeX{} is able, using well-defined
interfaces, to pass structure information to a backend, we expect that
support for other structured output formats will follow. Such work may
be undertaken by us or by other teams, possibly in parallel to later
phases of the project.


\subsection{Current status and achievements}

As mentioned earlier, \LaTeX{} was originally designed, as was
essential 40 years ago, to be very economical with computer resources;
the implementation therefore worked very hard to discard information
as soon as it was no longer needed for the compilation of a
document. For print output, which was all that was produced back then,
these discards included most of the structural information since this
was no longer useful once the visual representation had been
determined.
%
An important part of the early work on this project was therefore to
alter \LaTeX{}'s inner workings by adding code that preserves this
structural information from the source and adds it to the \PDF.

Another part of this early \enquote{background} work was to
standardize (and often to provide, for the first time) code interfaces
into which extension packages can safely hook.  The use of these
interfaces, rather than directly overwriting internal \LaTeX{}
functions (as was commonly done in the past), avoids the problem that
such packages would often break when used in certain combinations, or
break when \LaTeX{} internals changed.  Moreover, it means that these
packages can automatically benefit from the existence of extended
workflows (such as those which produce tagged \PDF).

Most of these interfaces are now in place in the \LaTeX{} kernel. What
remains (as a huge task) is to upgrade many of the core extension
packages so that they make use of the new functionalities; this will
enable the retirement of some of the existing code that directly
overwrites \LaTeX{} internals, or that makes assumptions (about those
internals) that will become invalid in the future.



The next large phase of the project was to provide automatic tagging
for a subset of \LaTeX{} documents. This task is largely finished and
therefore most documents that are restricted to using only the
commands and environments described in Leslie
Lamport's \enquote{\LaTeX{} Manual}~\cite{Lamport:1994} can be
automatically tagged by adding a single configuration line at the top
of the document.  We say \enquote{largely finished} because a few such
elements, or element combinations, are not yet covered at the time of
writing.

On the other hand, a number of extension packages that go beyond
Lamport are already supported, most importantly much of \pkg{amsmath}
(providing extended math capabilities) and \pkg{hyperref}
(enhancing \LaTeX{} with interactive hyperlinking features). Also
already supported are some of the major bibliography packages, such
as \pkg{natbib} and
\pkg{biblatex}.


The project is thus by now capable of producing \PDF~2.0 documents that
conform to the new \acro{PDF/UA-2} standard. In fact, after correcting a
small number of issues (not directly related to tagging) in the class
file for this conference we have been able to deliver this article as
a fully tagged \PDF~2.0 document.


\subsection{Ongoing and future project tasks}

At present, tagging support for the core document elements in a
\LaTeX{} document is still
at the prototype level, which means that it works for the
standard \LaTeX{} classes and for the document elements provided by
the \LaTeX{} kernel, but it may or may not work with extension
packages or classes that alter the implementation of these document
elements, or that provide completely new elements.

To make further progress, some of the interfaces for tagging will
first need to be finalized.  Then all major extension packages, as
well as all important third-party document classes, will need
analyzing and possibly updating.\footnote{The number of widely used
packages, e.g., those described in \titleref{The \LaTeX{} Companion,
third edition}~\cite{TLC3}, amounts to roughly 500, so this evaluation
and code adjustment forms a substantial part of the remaining project
work, and most likely will require additional volunteer support.}  The
tasks here are to identify all of the legacy low-level code for which
the kernel now provides tagging-aware replacements, and then, in
cooperation with their maintainers, to make the necessary updates to
all these packages and classes.

In addition, any packages and classes that provide new document
elements will need to specify how these elements are supposed to be
tagged.  Some interfaces already exist to help with this process, but
it is likely that most of these will require further refinement when
tested in the field.

The remaining phases of the project, as outlined in the
\enquote{Feasibility Evaluation Study}~\cite{feasibility-study},
cover further support for other \PDF\ standards, and an improved
interface to comprehensive metadata.
%
There are also a number of research problems that need to be solved in
order for authors to easily generate high-quality tagged \PDF\ documents
from their \LaTeX\ sources. These are outlined below.




\section{Specific aspects of the project work}

We now take a look at a few specific aspects of the project work that
are related to challenging problems and pose interesting research
questions. These topics are: the development of a more granular tag
set; the handling of formulas; the need for an extended table
specification syntax; and the handling of language and script related
requirements.


\subsection{The existing tag set support in \PDF}

When \PDF\ (already in version 1.3) first introduced a structure tree
into the format, to support the inclusion of the document's logical
structure, it used only a fairly minimal set of structure tags that
were largely modeled after the basic \HTML\ tag set.

For example, for mathematical formulas there was only
the \verb|<Formula>| tag itself, with no possibility to add further structure
within the formula.  For accessibility, all that was available was
an \enquote{alt} attribute in which one could add a textual
description of the formula's content.  In a similar manner, other
areas of document structures were (over)simplified in the tag set:
e.g., for all types of floating elements there is only the
tag \verb|<Aside>| that they must share with margin notes (and even
that tag is available only in \PDF~2.0). For code elements, whether
they are small snippets or long, commented listings, there is only a
single \verb|<Code>| tag, and there is no option to accurately
describe the handling of spaces and new lines within code listings.
There is a tag (again only in \PDF~2.0) to denote footnotes; but if the
document contains several types of structured (and possibly nested)
notes, then there is no way to adequately describe this without losing
possibly crucial information.


As is also the case with \HTML, the relationships between these tags
define a fairly simple document model that is not sufficiently rich,
so that it cannot express (or not correctly express) many real-life
documents; this is often due to the fact that certain elements appear
in such documents with nesting relationships that are not permitted by
the inclusion rules defined in \ISO\ 32005~\cite{pdfspec-iso32005_2023}.

All this means that, when preparing a \PDF\ to be \acro{PDF/UA-2} or
\acro{PDF/UA-1}
compliant, compromises have to be made and some of the structural
information may thus get lost.

As part of the project we are therefore developing an extended tag set
(currently called the \enquote{\LaTeX{} namespace}) that describes the
logical structure of (complex) documents in more granular detail; this
will help \PDF\ processors (such as viewers) that understand this
namespace to make better use of a document’s structure.  Ideas from
this development may also prove useful in conjunction with future
\HTML5 developments.





\subsection{The \LaTeX{} namespace}

\LaTeX{} is an open system that allows for structural extensions (and
even changes to structures) in every direction. It is therefore not
possible to define a fixed (definitive) document model that is both
valid and comprehensive for each and every conceivable \LaTeX{}
document.

However, it is possible to define a document model which captures the
majority of \LaTeX{} documents that are out there in the real
world. If this is combined with methods to extend (and possibly alter)
the document model whenever necessary for special structural
extensions or changes, we are confident that a comprehensive solution
can eventually be provided.

As part of the project we are therefore developing a \enquote{standard
namespace} that fully describes the \LaTeX{} document model (in the
sense outlined above).  This tag set will thus be noticeably more
detailed and comprehensive than those offered by \PDF~2.0 and \HTML5.
We are working with the \PDF\ Association~\cite{PDFA} and various
application producers to ensure that this namespace will, when
complete, become a recognized resource (preferably acknowledged in
future revisions of the
\PDF\ standard); it may also be more generally useful as an \XML\ schema.
This will, for example, allow \PDF\ and other applications to directly use
the extended
tag set it provides; and this will enable such applications to make
better use of the
information contained in the document, whether for accessibility
support or for other purposes.


For applications that do not (yet) understand this new namespace, we
provide role-mapping back into \PDF~2.0 (or \PDF~1.7) as necessary; but
of course, in that case the more granular information provided by the
tags in the new namespace will get at least partially lost.

Additionally, we will be providing interfaces that allow package or
class developers who extend the standard \LaTeX{} structures to
specify how their new commands or environments map into the \LaTeX{}
namespace (and from there, if necessary, are role-mapped back to the
\PDF\ tag set).



\subsection{Formulas in \acro{STEM} documents}


\LaTeX{} is well-known and appreciated for its ability to describe and
format mathematical or other formulas with a high degree of
flexibility and unsurpassed quality. This is one of the reasons why we
see a huge proportion of the documents in \acro{STEM} disciplines such as
mathematics, physics, and computer science being produced
using \LaTeX{}.

As described by Neil Soiffer in his keynote for the \acro{DEIMS}
2021 conference~\cite{Soiffer:deims2021}, there are basically three methods for
making such formulas accessible in a tagged \PDF. One option is to use
static text that can be attached as \enquote{alternative} natural
language text to the formula structure, which is then read by an
\enquote{assistive technology} (\acro{AT})
application. While rather easy to implement, this method has various
drawbacks: it does not allow for braille generation or for exploration
of the equation; and the text often must be hand-crafted to avoid
problems with reading software whose heuristic usually ignores certain
symbols such as punctuation or braces.


A second option is to add marked-content operators to the \PDF\ stream
and then build a \MathML\ structure tree that references this marked
content.  This method leads to a large
structure tree with many objects, since this tree will be very fine-grained.

There are a number of problems with this second
approach. It is difficult for
\LaTeX{} (without the help of a real programming language) to generate
correct and useful \MathML\ while building the \TeX{} math list.
Furthermore, when the still widely used \pdfTeX{} engine is producing
the \PDF, there is a high chance that the combined processes (of
simultaneously adding the necessary tagging-related material to the
content stream while formatting the formula) will alter the spacing
of the formula and thus render the visual representation
invalid. There may be technical solutions to circumvent these issues
and this is an area of active research. However, it is likely that
this option can only be implemented successfully if the \LuaTeX{}
engine is used, because then a suitable programming language (i.e.,
Lua) is available and, furthermore (because of extended functionality
in \LuaTeX{}), it becomes possible to delay adding the necessary extra
material to the content stream until after \LaTeX{} has completed the
formatting of the formula with the correct spacing.

The last option is to make use of so-called \enquote{associated files}
(\acro{AF}) that were introduced in \PDF~2.0: these are files directly
embedded into the \PDF\
that can be attached to a structure element.\footnote{Note that
\enquote{associated files} do not exist as
separate physical entities at the operating system level;
thus they are not in fact \enquote{files} in the normal sense of the word.}
%
Each such \enquote{embedded \acro{AF}} can
contain, for example, a \MathML\ representation for a formula, or
its \LaTeX{} source or some additional commentary text; more than one
of these can be attached to each structure.  The \acro{AF} approach is
simpler and easier to implement, and it also allows the use of \MathML\
representations that do not closely follow the visual output; but it
has the drawback that the \MathML\ in the \acro{AF} file is associated only to
the formula as a whole and it is therefore not possible to synchronize
parts of the \MathML\ representation with the corresponding parts of the
formula in the typeset document, as is necessary to support navigation
of formulas and highlighting them.  This method may require that the
\acro{AT} software overlays the printed output with its own rendering of the
\MathML\ (which may differ substantially from the original rendering).

The two last options, \MathML\ in the structure tree or in associated
files, both suffer from a lack of support in current \PDF\ viewers and
\acro{AT} software: Neil Soiffer's optimistic statement in 2021 that
\enquote{\textit{Adobe’s \API\ will likely incorporate this ability in the
future}} has not yet come true.

Because of this, \LaTeX{} currently follows a three-fold strategy in
the prototype for math tagging: it incorporates the \LaTeX{} source as
alternate text for the formula, under the assumption that the \LaTeX{}
syntax is understandable to most readers of mathematics; and it also embeds
the \LaTeX{} source as an associated file.
%
%Additionally, for all (or
%a selection of) the formulas, MathML representations can be provided
%in an external file that are then also embedded as associated files.
Additionally, an external file can be constructed in which, for all (or
a selection of) the formulas, a \MathML\ representation is provided
that can be embedded in the \PDF\ as associated files.
%
Such an external file can be created, for example, with the help
of \texttt{tex4ht}\label{mml-external} or with the \LuaTeX{} engine.
At this point in time there is no fully automatic workflow implemented
for this, but with only a few adjustments it was already possible to
add \MathML\ associated files to all the formulas in the \pkg{amsmath}
user documentation.

The form of the final solution for formulas, and whether or not it is
necessary to offer customizable alternatives\Dash to cater for different
reader deficiencies or different user preferences\Dash are questions that
need active research to understand how to best serve consumers given
the currently limited functionality of \acro{AT} tools with respect to
\enquote{associated files}, etc.

There is another aspect of common \LaTeX{} usage that affects all three
of these
method, and is also found in many other areas beyond formulas and
\acro{STEM}: the inclination of authors to invent new symbols, notation
systems, and command names. This is nowadays exacerbated by the
widespread failures to take accessibility into account.  Such ad hoc
extensions make it difficult to fully automate any tagging process.
Overcoming this will need both technical support for such extensions
and also, perhaps more importantly, encouragement of authors to keep
accessibility in mind when writing documents.

The approach that will most likely be adopted to deal with this issue
is as follows: by default, assume that new commands are simply
abbreviations and that, by recursively replacing each of these with its
definition, we eventually get to something that can be automatically
tagged by using standard methods. For cases where this does not work,
there will be interfaces with which the author of the document (or the
package developer, if the command is defined there) can specify how
the command should be interpreted when providing tagged output (e.g.,
\MathML).




\subsection{Extensions to \LaTeX{}'s table handling}

In most cases, the \LaTeX{} source will contain all the necessary
information about the logical structure of a document, so that it is
possible to automatically transform the source into richly tagged \PDF\
output. There is one noticeable exception: \LaTeX{}'s handling of
tabular data. This arises since standard \LaTeX{}, and most extension packages, do not
describe table data through structural information; rather, they do
this in a purely visual fashion, describing only the content that
should go into each cell.  Thus no information is supplied concerning
important relationships between cells, such as which are the header or
sub-header cells, or to which cells some header cell applies.

Thus, while it is fairly trivial to tag tables as simply consisting of
table rows and table data cells, determining the header cells can be
done only by the use of heuristics (e.g., cell formatting changes done
through \verb|\multicolumn| are likely to represent header cells, or
certain rules in a table may indicate header rows).  However, any such
heuristic will have a noticeable number of counterexamples.

It is therefore an important task to develop good heuristics that
correctly cover a large proportion of the tables in legacy documents;
and in addition to develop a syntax extension for \LaTeX{} that
allows authors to specify such logical structure explicitly in case
the heuristics fail or they wish to specify explicitly the logical
structure of the table. This syntax extension has to be done in a
lightweight way, i.e., without putting an unnecessary burden onto
the authors. Furthermore, it should be upwardly compatible with the
existing syntax so that it is easily possible to enhance documents
with only small alterations to the original source.

It is also important to develop methods that enable authors to easily check
\LaTeX{}'s interpretation of the logical structure of a table without the
need to examine the final \PDF, so that they can overwrite the
heuristics when necessary. This is an area of active research.


\subsection[Support for different scripts and\texorpdfstring{\\}{} languages]
           {Support for different scripts and languages}


Historically, the \TeX{} engine and \LaTeX{} were developed for
\ASCII-based, English documents and then (with \TeX~3.0) extended to
support other languages and scripts\Dash at first, because of the
restrictions to 8-bit codepages, mostly for languages using Latin
scripts, but later also to non-Latin scripts, such as Greek or
Cyrillic, as well as more diverse scripts. Initially, the solutions
for all such scripts required complex font setups (as done, e.g., by
the \pkg{CJK} package), special processors to handle transliterations,
and engine extensions to handle, for example, right-to-left scripts or
special input encodings.

The advent of Unicode and the Unicode-aware engines
(\XeTeX{}, \LuaTeX{} and \upTeX{}) led to the existence of simpler,
and much more powerful, setups; therefore, most scripts are now well
supported in \LaTeX{}\Dash perhaps with the exception of scripts that
change the writing direction, since this isn't part of the
original \LaTeX{} design and thus often requires overwriting many
standard commands.

The project currently concentrates on documents that use Latin scripts
or scripts with similar characteristics.  The correct tagging
conventions to use with other types of script are not yet known by us:
e.g., how to deal with direction changes or ruby characters.
%
When using scripts (such as Latin) that typically
use \enquote{whitespace} to delimit \enquote{words}, tagged \PDF\ has a
requirement that even within the typeset content stream these words
must remain delimited by an explicit \enquote{whitespace
character}~\cite[§14.8.2.6.2]{pdfspec-iso32000-2_2020}.  This
conflicts with the normal practice of \TeX{} typesetting engines since
they do not naturally add such delimiter characters; however,
both \pdfTeX{} and \LuaTeX{} have been modified to provide workarounds
for this.\footnote{Engines, such as \XeTeX{}, that do not offer this
workaround can therefore not be used to produce \acro{PDF/UA} documents
involving scripts that separate words with spaces.}
%
We also do not yet know to what extent the many external packages
supporting diverse scripts and languages will need to be adapted for
the support of tagging.  To research these topics, help from users and
developers with in-depth knowledge of such scripts will be needed.




\section{Some \TeX{}nical details}

In this final section we take a brief look at two technical aspects of
the project work.  Both will be covered in more depth during the
demonstration session at the conference.

The first subsection explains how to set up a document (such as this
one) so that it will automatically produce tagged \PDF.
This should, we hope, enable you to immediately experiment with the
addition of such tags to your own works. If you want to provide feedback
on any issues that you encounter, or to provide suggestions for
improvements,
%these should be added
we suggest adding them
to the project repository
\url{https://github.com/latex3/tagging-project}, using either the \texttt{issues} or
the \texttt{discussions} page, as appropriate.

This is followed by some background information on the experiments we
are currently conducting to automatically include in the \PDF\ \MathML\
representations of all the formulas in a document. We expect this
work to become available for public testing during 2024/Q2.


\subsection{How to enable tagging}


Until recently there was no dedicated location in \LaTeX{} documents
to declare settings that affect the document as a whole. Settings had to
be placed somewhere in the preamble or as class options, or sometimes
even as package options. For some such settings this was
problematic, e.g., setting the \PDF\ version is only possible if
the \PDF\ output file has not yet been opened, which can be caused by
loading one or another package.
%
For the \enquote{\LaTeX{} Tagged \PDF\
project}~\cite[p.\,17]{feasibility-study} further metadata about the
whole document (and its processing) needs to be specified, and again
all this data should be placed in a single well-defined place.

For this reason we introduced (in June 2022) the command
\verb/\DocumentMetadata/ so as to unify all such settings in one place.
This command takes one argument that should contain a key/value list
specifying all the document metadata for the current
document.\footnote{At this point in time only a few keys are accepted,
e.g., to set the \PDF\ version, the language, a \PDF\ standard and to load
a color profile.}  This should be placed at the very beginning of the
document, i.e., \emph{before}
\verb/\documentclass/; it will produce an error if found later.

\begin{figure*}\small
\begin{verbatim}
         <div>
         <h2>\mml 65</h2>
         <p>\begin{math}\sqrt [\beta ]{k}\end{math}</p>
         <p>656E4D3BB4F29D20A1B2CBCB35C35E7E</p>
         <math xmlns="http://www.w3.org/1998/Math/MathML" display="inline">
         <mroot>
         <mi>k</mi>
         <mi>&#x03b2;</mi>
         </mroot>
         </math>
         </div>
\end{verbatim}
\vspace{-.5\baselineskip}
\caption{Sample entry with MathML data for associated file (AF)}
\label{fig:mml}
\end{figure*}

The
\verb/\DocumentMetadata/ command also loads the \LaTeX\ \PDF\ management bundle,
which provides various \PDF-related commands that are needed to create
a tagged \PDF.
%
It
%The \verb/\DocumentMetadata/ command
also accepts
the \texttt{testphase} key, which is of a temporary nature since it is
needed only while new functionality is being introduced for
testing. This key is used to load specific tagging support: this
article, for example, uses the following:
\begin{verbatim}
  \DocumentMetadata{
     testphase={phase-III,table},
     pdfversion=2.0,
     pdfstandard=a-4,
  }
\end{verbatim}
which loads the tagging support from \texttt{phase-III} (basic
document elements) and \texttt{table} (newly developed prototype code
for tagging of \env{tabular}-like environments not yet integrated in
any test phase).
%
In addition, the \PDF\ version is set to 2.0 and it specifies that the
\PDF\ should be compliant with the \acro{PDF/A-4} standard. This is all
that was necessary to produce the tagged version of this
article.\footnote{This paper does not contain any tabular material,
thus \texttt{table} is actually unnecessary for tagging this article.
The setting was added to show how the interface can be used when new
functionality is made available.}


Eventually, the testphase code will move (once all components are considered stable)
into the \LaTeX{} kernel itself and the \texttt{testphase} key will
vanish.  Tagging will continue to require a \verb/\DocumentMetadata/
declaration, but will then use a simple \texttt{tagged=true} key (name to
be decided).




\subsection{Inclusion of external \MathML}


As outlined in section~\ref{mml-external} on
page~\pageref{mml-external} we are currently experimenting with a
scheme in which externally provided \MathML\ is embedded in the \PDF\
as \acro{AF}s. The \MathML\ for the formulas is provided in an external
file containing one or more \verb=\mml= commands with the
format shown in figure~\ref{fig:mml} (i.e., surrounded by \HTML\ tags so that it can be proofread in a browser).

The first argument to \verb=\mml= is a label (e.g., a number) to
that uniquely identifies this \MathML\ snippet; the second argument
contains the \LaTeX{} source for the \MathML.
%
The third argument is the \acro{MD5} hash of the \LaTeX{} source.
Its use ensures that the \PDF\ file will contain
only one \acro{AF} for any
formula, even if a formula is repeated several times: for example, if the
\LaTeX{} source document repeatedly uses \verb=$\beta$=, then each repetition
of exactly this formula will reference the same embedded \acro{AF}, which
means that the \acro{PDF} file does not become
unnecessarily large.\looseness-1

The final argument contains the corresponding \MathML.  In our current
experiments the \MathML\ is generated from the \LaTeX{} source by
processing it with \texttt{tex4ht}, with some further processing to add the
\acro{MD5} hash values and with some manual corrections to improve the resulting
\MathML. One advantage of using an external file at this stage is to
allow the \MathML\ to be validated before being embedded as an \acro{AF} in the
\PDF. The \MathML\ could potentially be generated by other  \TeX\ to \MathML\
conversion programs such
as \texttt{latexml} or \texttt{luamml}, which would allow
experimentation with different pipelines to construct
associated files containing \MathML.


This file is then input at the beginning of the document and each
\MathML\ (with a unique hash value) is embedded in the \PDF\ as the
content stream of an \acro{AF}.
%
%The \LaTeX{} code to produce tagged PDF
%then checks, for each math formula in the document, whether a MathML
%file with the same MD5 hash already exists; and if one does then its
%AF is simply referenced in the structure.
%
The \LaTeX{} code to produce tagged \PDF\ then checks, for each math
formula in the document, whether an associated file containing  \MathML\ for this
formula has already been added to the \PDF, and, if so, a reference to this
\MathML\ associated file is added to the \verb|<Formula>| structure element being
constructed.  Therefore, if the same math expression occurs more than
once (as a complete formula) then each occurrence will reference this
same \MathML{}~\acro{AF}.\looseness=1


Currently, the generation of the file of \MathML\ fragments requires some
manual editing and explicit execution of conversion programs. The next
step will be to create scripts that will: run directly in a \LuaTeX{}
compilation; fully automate the generation of the \MathML\ fragments
from the \LaTeX{} source; and validate this output.

\bibliographystyle{plain}
\let\url\tbsurl
\SetBibJustification{\raggedright\advance\itemsep by 1pt}
\small

\bibliography{tb139mitt-deims24-bibtex}

\makesignature

\end{document}
